Supplementary Information

### 2021-05-19 doc/manuscript/ForestGEO_treerings.Rmd
work on intro

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and year.
In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019]. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g.,  @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010] integrate effects of precipitation and temperature, soil moisture and atmospheric vapor pressure deficit (VPD) affect plant growth differently [@novick_increasing_2016], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

>>>>

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
Specifically in the case of drought, soil moisture (shaped by precipitation and evapotranspiration) and atmospheric vapor pressure deficit (VPD; driven by temperature) affect plant growth differently [@novick_increasing_2016; @humphrey_soil_2021], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
